Appendix

### 2025-12-19 manuscript_rmarkdown/manuscript_appendix.tex
Update manuscript_appendix.tex

\title{Appendix}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-10-29}

@@

\section{Appendix C: Empirical central normality test critical
values}\label{appendix-c}

@@

In the main manuscript, transformation methods are compared by ranking
within each experiment, defined by the combination of sequence and
absence/presence of outliers. This allows for direct comparison of
methods between different experiments.

Table \ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
shows the average normalised residual error for these transformation
methods. This metric is computed from the residual error and the
residual error of the central portion (\(\kappa = 0.80\)) for sequences
without and with outliers, respectively. In each experiment, the
respective residual error was normalised by dividing with the sequence
length, and can be interpreted as the average residual error of a single
sample. This reduces the effect of sequence length on the residual
error.

Table
\ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
shows the average normalised difference in residual error between the
best method for each experiment and the individual methods. As above,
normalisation was done by division by the length of the sequence
corresponding to each experiment.

@@

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. Residual errors were normalised by sequence length to reduce the effect of sequence length on the residual error.
The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
\begin{tabular}{l | l r r r r r r}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average difference in residual error between Yeo-Johnson
transformation methods and the best transformation method within each
experiment, normalised by the corresponding sequence length. This is based on
either the residual error (without outliers) or the residual error of the
central portion (with outliers; $\kappa = 0.80$) over 3 datasets with 100000
sequences each. Residual errors were normalised by sequence length to reduce the
effect of sequence length on the residual error. The clean dataset consists of
sequences derived through inverse Yeo-Johnson transformation of data sampled
from a standard normal distribution. The dirty dataset contains sequences
sampled from asymmetric generalised normal distributions, centred at 0. The
shifted dataset also contains sequences sampled from asymmetric generalised
normal distributions, but centred at 100, and scaled by 0.001. Several
transformation methods include normalisation before transformation, indicated by
z-score normalisation (norm.) or robust scaling. A lower difference is better,
with a minimum value of 0.0. Discrepancies with the rank-based method may occur,
as normalised differences in residual error are not fully comparable between
experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
\begin{tabular}{l | l r r r r r r}

@@

\subsection{Ranking Yeo-Johnson transformations: effect of sequence
length}\label{ranking-yeo-johnson-transformations-effect-of-sequence-length}

The main manuscript shows the overall ranking of different
Yeo-Johnson-based power transformation methods across three sets of
problems. To assess the effect of limited sequence lengths, we perform
ranking on subsets of the experiment. The results for ranking the subset
of sequences with 30 samples or fewer (\(n = 16001\)) is shown in Table
\ref{tab:comparison_methods_simulations_yeo_johnson_10_30}. In the clean
dataset without outliers, where all sequences were sampled from a normal
distribution and then underwent an inverse Yeo-Johnson transformation,
conventional and location- and scale-invariant are ranked similarly,
with no clear advantage to any method. For the remaining datasets, the
robust location- and scale-invariant method ranked best. It slightly
outperformed (robustly scaled) conventional Yeo-Johnson transformations
and (z-score normalised) robust transformations
\citep{Raymaekers2024-zf}, in datasets without and with outliers,
respectively.

In the subset for larger sequence lengths, with at least 1000 samples
per sequence (\(n = 33037\)), the cost of making the location- and
scale-invariant transformation robust against outliers becomes more
apparent (Table
\ref{tab:comparison_methods_simulations_yeo_johnson_1000_10000}). In the
clean dataset without outliers, down-weighting of the tails of the
distribution by the robust location- and scale-invariant method leads to
larger residuals compared to other methods (Table
\ref{tab:comparison_methods_simulations_yeo_johnson_1000_10000_avg_residual}.

@@

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 16001 sequences each. Residual errors were normalised by sequence length to reduce the effect of sequence length on the residual error.
Each sequence contained 30 samples or fewer. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_10_30_avg_residual}
\begin{tabular}{l | l r r r r r r}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 33037 sequences each. Each sequence contained between 1000 and 10000 samples.
The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson_1000_10000}
\begin{tabular}{l | l r r r r r r}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 16001 sequences each. Residual errors were normalised by sequence length to reduce the effect of sequence length on the residual error.
Each sequence contained between 1000 and 10000 samples. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_1000_10000_avg_residual}
\begin{tabular}{l | l r r r r r r}

@@

Robust transformations are hypothesised to have a cost in efficiency for
data without outliers, i.e.~clean data. Here we draw nine sequences with
elements randomly drawn from a standard normal distribution
\(\mathcal{N}(0,1)\). Subsequently, we perform an inverse transformation
\(\left(\phi^{\lambda, 0, 1}\right)^{-1}\), with
\(\lambda \in \left(0, 2\right)\).

The sequences drawn resulted from the permutations of the number of
elements of each sequence (\(n \in \{30, 100, 500\}\)) and
transformation parameter for the inverse transformation
\(\lambda \in \{0.1, 1.0, 1.9\}\). Each sequence then underwent power
transformation using Box-Cox and Yeo-Johnson transformations. For
Box-Cox transformations, each sequence was shifted prior to inverse
transformation ensuring all elements were strictly positive after
inverse transformation.

>>>>

\title{Appendix}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-12-19}

@@

\section{Appendix C: Empirical central normality test critical
values}\label{appendix-c-empirical-central-normality-test-critical-values}

@@

In the main manuscript, transformation methods are compared by ranking
within each experiment, defined by the combination of feature and
absence/presence of outliers. This allows for direct comparison of
methods between different experiments.

Table \ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
shows the average normalised residual error for these transformation
methods. This metric is computed from the residual error and the
residual error of the central portion (\(\kappa = 0.80\)) for features
without and with outliers, respectively. In each experiment, the
respective residual error was normalised by dividing with the feature
length, and can be interpreted as the average residual error of a single
sample. This reduces the effect of feature length on the residual error.

Table
\ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
shows the average normalised difference in residual error between the
best method for each experiment and the individual methods. As above,
normalisation was done by division by the length of the feature
corresponding to each experiment.

@@

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 features each. Residual errors were normalised by feature length to reduce the effect of feature length on the residual error.
The clean dataset consists of features derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains features sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains features sampled from asymmetric generalised normal distributions, but centred at 1000.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors 
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
\begin{tabular}{l | l r r r r r r}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average difference in residual error between Yeo-Johnson
transformation methods and the best transformation method within each
experiment, normalised by the corresponding feature length. This is based on
either the residual error (without outliers) or the residual error of the
central portion (with outliers; $\kappa = 0.80$) over 3 datasets with 100000
features each. Residual errors were normalised by feature length to reduce the
effect of feature length on the residual error. The clean dataset consists of
features derived through inverse Yeo-Johnson transformation of data sampled
from a standard normal distribution. The dirty dataset contains features
sampled from asymmetric generalised normal distributions, centred at 0. The
shifted dataset also contains features sampled from asymmetric generalised
normal distributions, but centred at 1000. Several
transformation methods include normalisation before transformation, indicated by
z-score normalisation (norm.) or robust scaling. A lower difference is better,
with a minimum value of 0.0. Discrepancies with the rank-based method may occur,
as normalised differences in residual error are not fully comparable between
experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
\begin{tabular}{l | l r r r r r r}

@@

\subsection{Ranking Yeo-Johnson transformations: effect of feature
length}\label{ranking-yeo-johnson-transformations-effect-of-feature-length}

The main manuscript shows the overall ranking of different
Yeo-Johnson-based power transformation methods across three sets of
problems. To assess the effect of limited feature lengths, we perform
ranking on subsets of the experiment. The results for ranking the subset
of features with 30 samples or fewer (\(n = 16001\)) is shown in Table
\ref{tab:comparison_methods_simulations_yeo_johnson_10_30}. In the clean
dataset without outliers, where all features were sampled from a normal
distribution and then underwent an inverse Yeo-Johnson transformation,
conventional and location- and scale-invariant are ranked similarly,
with no clear advantage to any method. For the remaining datasets, the
robust location- and scale-invariant method ranked best. It slightly
outperformed (robustly scaled) conventional Yeo-Johnson transformations
and (z-score normalised) robust transformations
\citep{Raymaekers2024-zf}, in datasets without and with outliers,
respectively.

In the subset for larger feature lengths, with at least 1000 samples per
feature (\(n = 33037\)), the cost of making the location- and
scale-invariant transformation robust against outliers becomes more
apparent (Table
\ref{tab:comparison_methods_simulations_yeo_johnson_1000_10000}). In the
clean dataset without outliers, down-weighting of the tails of the
distribution by the robust location- and scale-invariant method leads to
larger residuals compared to other methods (Table
\ref{tab:comparison_methods_simulations_yeo_johnson_1000_10000_avg_residual}.

@@

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 16001 features each. Residual errors were normalised by feature length to reduce the effect of feature length on the residual error.
Each feature contained 30 samples or fewer. The clean dataset consists of features derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains features sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains features sampled from asymmetric generalised normal distributions, but centred at 1000.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors 
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_10_30_avg_residual}
\begin{tabular}{l | l r r r r r r}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 33037 features each. Each feature contained between 1000 and 10000 samples.
The clean dataset consists of features derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains features sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains features sampled from asymmetric generalised normal distributions, but centred at 1000.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson_1000_10000}
\begin{tabular}{l | l r r r r r r}

@@

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 16001 features each. Residual errors were normalised by feature length to reduce the effect of feature length on the residual error.
Each feature contained between 1000 and 10000 samples. The clean dataset consists of features derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains features sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains features sampled from asymmetric generalised normal distributions, but centred at 1000.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors 
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_1000_10000_avg_residual}
\begin{tabular}{l | l r r r r r r}

@@

Robust transformations are hypothesised to have a cost in efficiency for
data without outliers, i.e.~clean data. Here we draw nine features with
elements randomly drawn from a standard normal distribution
\(\mathcal{N}(0,1)\). Subsequently, we perform an inverse transformation
\(\left(\phi^{\lambda, 0, 1}\right)^{-1}\), with
\(\lambda \in \left(0, 2\right)\).

The features drawn resulted from the permutations of the number of
elements of each feature (\(n \in \{30, 100, 500\}\)) and transformation
parameter for the inverse transformation
\(\lambda \in \{0.1, 1.0, 1.9\}\). Each feature then underwent power
transformation using Box-Cox and Yeo-Johnson transformations. For
Box-Cox transformations, each feature was shifted prior to inverse
transformation ensuring all elements were strictly positive after
inverse transformation.
